Flexible Probabilistic Forecasting with Distributional Trees and Forests

### 2018-03-27 pkg/disttree/inst/MIP-Seminar/slides.Rnw
added observations
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2097 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: b88abb1ae358e46429b39966d98827ce03aa5df4

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.75\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit a global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
Estimate model parameters $\hat{\theta}$.
\item<5-> Evaluate how well the model fits the data (for each parameter and each observation).
\item<7-> Choose the covariate $X$ with the strongest influence or association to
goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.2\textwidth}
\vspace{-0.5cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$Y$};
}
\visible<4->{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$\mathcal{D}(Y;\hat{\theta}$)};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
\draw[-, line width=0.8pt] (n0) -- (n1);
\draw[-, line width=0.8pt] (n0) -- (n2);
}
\visible<9>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<10->{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
\begin{adjustwidth}{-0.0em}{-3em}
\begin{enumerate}
\setcounter{enumi}{3}
\item<8-> Find the split point $p$ which leads to the highest improvement.
\item<11-> Repeat steps 1--4 recursively in the subgroups
    until some stopping criterion is met.
\end{enumerate}
\end{adjustwidth}
\vspace{0.4cm}
\visible<12->{
{\bf Forest:} Ensemble of trees.
\begin{itemize}
\item Varying subsamples.
\item Combine/average over trees.
\end{itemize}
}
\end{frame}

@@

getZoo <- function( x, variable ) {
  x <- subset( x, varname == variable )
  if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
  # Else create zoo
  x <- zoo( subset(x,select=-c(varname,timestamp,step)),
            as.POSIXct(x$timestamp,origin="1970-01-01") )
  x
}
GFS_t2m  <- getZoo( GFS, "tmp2m" )
# temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
# => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
GFS_t2m <- GFS_t2m + 10.7

GFS_rain <- getZoo( GFS, "apcpsfc" )
# precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
GFS_rain24 <- GFS_rain[c((1:10)*4),]
for(i in 1:NROW(GFS_rain24)){
  GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
}
# first value of rain is 18 hours later than temperature 
#(both start at 6:00 but rain is cumulated over 24 hours)
# -> drop first 3 values of temperature
GFS_t2m <- GFS_t2m[-c(1:3),]

@@

>>>>

\begin{frame}
\frametitle{Building Distributional Trees and Forests}
\begin{minipage}{0.75\textwidth}
{\bf Tree:}
\begin{enumerate}
\item<3-> Fit a global distributional model $\mathcal{D}(Y; \theta)$: \\ % to the whole data set:\\
Estimate model parameters $\hat{\theta}$.
\item<5-> Evaluate how well the model fits the data (for each parameter and each observation).
\item<7-> Choose the covariate $X$ with the strongest influence or association to
goodness-of-fit of $\mathcal{D}(Y; \theta)$ as split variable.
\end{enumerate}
\end{minipage}
\begin{minipage}{0.2\textwidth}
\vspace{-0.5cm}
\begin{tikzpicture}
\visible<2-3>{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$Y$};
}
\visible<4->{
\node[ellipse, fill=HighlightBlue!70, align=center, scale = 0.7, minimum width=60pt, minimum height = 30pt] (n0) at (0.8, 2) {$\mathcal{D}(Y;\hat{\theta}$)};
}
\visible<6-8>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {?};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {?};
\draw[-, line width=0.8pt] (n0) -- (n1);
\draw[-, line width=0.8pt] (n0) -- (n2);
}
\visible<9>{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$Y_1$};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$Y_2$};
\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\visible<10->{
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n1) at (0, 0) {$\mathcal{D}(Y_1;\hat{\theta_1}$)};
\node[rectangle, fill=HighlightOrange!70, align=center, scale = 0.7, minimum width=50pt, minimum height = 20pt] (n2) at (1.6, 0) {$\mathcal{D}(Y_2;\hat{\theta_2}$)};
\draw[-, line width=0.8pt] (n0) -- (n1) node [midway, left] {\scriptsize $X \leq p$};
\draw[-, line width=0.8pt] (n0) -- (n2) node [midway, right] {\scriptsize $X > p$};
}
\end{tikzpicture}
\end{minipage}
\vspace{0.1cm}
\begin{adjustwidth}{-0.0em}{-3em}
\begin{enumerate}
\setcounter{enumi}{3}
\item<8-> Find the split point $p$ which leads to the highest improvement.
\item<11-> Repeat steps 1--4 recursively in the subgroups
    until some stopping criterion is met.
\end{enumerate}
\end{adjustwidth}
\vspace{0.4cm}
\visible<12->{
{\bf Forest:} Ensemble of trees.
\begin{itemize}
\item Varying subsamples.
\item Combine / average over trees.
\end{itemize}
}
\end{frame}

@@

# observations
load("STAGEobs_tawes_11121.rda")
# "2018-03-14 01:00:00 CET" in line 321
obs <- obs[-c(1:320),]

getZoo <- function( x, variable ) {
  x <- subset( x, varname == variable )
  if ( nrow(x) == 0 ) stop("Ups, variable seems not to exist at all!")
  # Else create zoo
  x <- zoo( subset(x,select=-c(varname,timestamp,step)),
            as.POSIXct(x$timestamp,origin="1970-01-01") )
  x
}
GFS_t2m  <- getZoo( GFS, "tmp2m" )
# temperature bias: model simulates temperature for height 1675 (~1070 m higher than true height)
# => add 10.7 degrees to account for bias (approx. 1 degree / 100 m)
GFS_t2m <- GFS_t2m + 10.7
# first value of rain is 18 hours later than temperature 
#(both start at 6:00 but rain is cumulated over 24 hours)
# -> drop first 3 values of temperature
GFS_t2m <- GFS_t2m[-c(1:3),]
GFS_t2m$obs <- obs[,"obs$tl"]
GFS_t2m <- GFS_t2m[c(1:40),]

GFS_rain <- getZoo( GFS, "apcpsfc" )
GFS_rain$obs <- obs[,"rr6"]
# precipitaion sums over 24 hours (00:00 UTC - 00:00 UTC +1day)
GFS_rain24 <- GFS_rain[c((1:10)*4),]
for(i in 1:NROW(GFS_rain24)){
  GFS_rain24[i,] <- colSums(GFS_rain[c((4*i-3):(4*i)),])
}
## FIX ME: add observations

@@

\begin{frame}[fragile]
\frametitle{Probabilistic Forecasting of Precipitation}
\vspace{-0.3cm}
\setkeys{Gin}{width=1.1\linewidth}
<<ensemble_1_plot+obs, fig=TRUE, echo=FALSE, width=9>>=
par(mar=c(0.4,0,0.2,0), oma=c(6,3.2,5,4))
layout( matrix(1:2,ncol=1) )
plot( GFS_t2m[,c(1,22)], screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
#plot( GFS_t2m_mean, screen=1, xaxs="i", xaxt="n", ylim = range(GFS_t2m)*1.05)
abline( v = main, lty = 2 )
mtext( side = 3, line = 1, cex = 1.2, font = 2, title )
mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Temperature [°C]" )
plot( GFS_rain24[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain24)*c(0,1.05), type = "o" )
#plot( GFS_rain[,1], screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
#plot( GFS_rain_mean, screen=1, xaxs="i", xaxt="n", yaxs="i", ylim=range(GFS_rain)*c(0,1.05) )
abline( v = main, lty = 2 )
mtext( side = 2, line = 2.3, cex = 1, font = 1, text = "Rain [mm/6h]" )
